---
title: "dyrk1a snRNA-seq — median+5SD filter, SCTransform, clustree"
author: "Sicong"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 9,
  fig.height = 6
)
```

## How to run

1. Matrices under `data/<sample_folder>/`.
2. QC: **nFeature min = 200**, **max = per-sample median + 5 × SD**, **mt_max = 3**, singlets only.
3. Step 6: merged filtered → SCTransform → Harmony; **LogNormalize fallback** if SCT fails.
4. **Step 6:** 6a SCT/Harmony → 6b clustree → 6c clustering plots (enable 6b/6c flags after previous step finishes).

Helpers: `scripts/snRNAseq_qc_filter_helpers.R`, `snRNAseq_median5sd_clustree_helpers.R`.

Set working directory to repo root (`snRNAseq_median5sd_clustering/`) in RStudio before running.

---

## Pipeline options

```{r pipeline-options}
if (!exists("find_repo_root", mode = "function")) {
  rr <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  for (i in seq_len(6)) {
    f <- file.path(rr, "scripts", "repo_root.R")
    if (file.exists(f)) {
      source(f)
      break
    }
    rr <- dirname(rr)
  }
}
repo_root <- find_repo_root()

run_step1_3_build_merge  <- TRUE
run_step4_qc_exploratory <- TRUE
run_step5_filter         <- TRUE
run_step6a_sctransform   <- TRUE
run_step6b_clustree      <- FALSE
run_step6c_clustering    <- FALSE

run_step4a_filter_diagnostics <- TRUE
run_step4b_round_grid_preview <- TRUE

nFeature_min   <- 200
sd_multiplier  <- 5
mt_max         <- 3

clustree_resolutions <- seq(0, 1, 0.1)
clustering_resolutions <- c(0.6)
step6_force_rerun <- FALSE

density_xlim_nFeature <- c(0, 6000)
density_xlim_nCount   <- c(0, 15000)

qc_zoom_y_min <- NULL
qc_zoom_y_max <- NULL
```

---

## Paths and sample map

```{r paths}
data_root <- file.path(repo_root, "dyrk1a_snRNAseq", "data")
out_dir   <- file.path(repo_root, "dyrk1a_snRNAseq", "out")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

gene_label      <- "dyrk1a"
merge_project   <- "dyrk1a_snRNA"
object_prefix   <- "dyrk1a"
merged_raw_rds  <- file.path(out_dir, "dyrk1a_merged_raw.rds")
merged_filt_rds <- file.path(out_dir, "dyrk1a_merged_filtered_median5sd.rds")
list_rds        <- file.path(out_dir, "dyrk1a_seurat_objs_snRNA_seq_list_with_scDblFinder.rds")

samples <- c(
  dyrk1a_wt_rep1  = "dyrk1a_WT1_DRT",
  dyrk1a_wt_rep2  = "dyrk1a_WT2_DRT",
  dyrk1a_wt_rep3  = "dyrk1a_WT3_DRT",
  dyrk1a_het_rep1 = "dyrk1a_HET1_DRT",
  dyrk1a_het_rep2 = "dyrk1a_HET2_DRT",
  dyrk1a_het_rep3 = "dyrk1a_HET3_DRT",
  dyrk1a_hom_rep1 = "dyrk1a_HOM1_DRT",
  dyrk1a_hom_rep2 = "dyrk1a_HOM2_DRT",
  dyrk1a_hom_rep3 = "dyrk1a_HOM3_DRT"
)

resolve_10x_dir <- function(sample_folder) {
  sample_dir <- file.path(data_root, sample_folder)
  required <- c("barcodes.tsv.gz", "features.tsv.gz", "matrix.mtx.gz")
  missing <- required[!file.exists(file.path(sample_dir, required))]
  if (length(missing) > 0) {
    stop("Missing 10x files in ", sample_dir, ": ", paste(missing, collapse = ", "))
  }
  sample_dir
}

paths_check <- data.frame(
  sample_id = names(samples),
  folder = unname(samples),
  matrix_dir = vapply(unname(samples), resolve_10x_dir, character(1)),
  stringsAsFactors = FALSE
)
paths_check
```

```{r source-helpers}
source(file.path(repo_root, "scripts", "snRNAseq_qc_filter_helpers.R"))
source(file.path(repo_root, "scripts", "snRNAseq_median5sd_clustree_helpers.R"))
```

---

## Packages

```{r install-packages, eval=FALSE}
install.packages(c("Seurat", "dplyr", "ggplot2", "tibble", "patchwork", "harmony", "future", "clustree"))
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("SingleCellExperiment", "scDblFinder", "glmGamPoi"), update = FALSE, ask = FALSE)
```

```{r packages}
suppressPackageStartupMessages({
  library(Seurat)
  library(dplyr)
  library(ggplot2)
  library(tibble)
  library(SingleCellExperiment)
  library(scDblFinder)
})
```

---

## Step 1 — Per-sample Seurat objects

```{r step1-create-objects, eval=run_step1_3_build_merge}
seurat_objs <- lapply(names(samples), function(sample_id) {
  folder <- samples[[sample_id]]
  tenx_dir <- resolve_10x_dir(folder)
  message("Reading ", sample_id, " from ", tenx_dir)
  counts <- Read10X(tenx_dir)
  obj <- CreateSeuratObject(counts = counts, project = sample_id)
  obj$orig.ident <- sample_id
  obj$genotype <- dplyr::case_when(
    grepl("_wt_", sample_id) ~ "WT",
    grepl("_het_", sample_id) ~ "HET",
    grepl("_hom_", sample_id) ~ "HOM",
    TRUE ~ NA_character_
  )
  saveRDS(obj, file.path(out_dir, paste0("seurat.obj_", sample_id, ".RNA.rds")))
  obj
})
names(seurat_objs) <- names(samples)

tibble(
  sample = names(seurat_objs),
  n_cells = vapply(seurat_objs, ncol, numeric(1))
)
```

---

## Step 2 — scDblFinder

```{r step2-doublets, eval=run_step1_3_build_merge}
for (sample_id in names(seurat_objs)) {
  message("scDblFinder: ", sample_id)
  sce <- as.SingleCellExperiment(seurat_objs[[sample_id]])
  sce <- scDblFinder(sce)
  seurat_objs[[sample_id]]$scDblFinder.score <- sce$scDblFinder.score
  seurat_objs[[sample_id]]$scDblFinder.class <- sce$scDblFinder.class
}

saveRDS(seurat_objs, list_rds)

doublet_summary <- bind_rows(lapply(names(seurat_objs), function(sample_id) {
  md <- seurat_objs[[sample_id]]@meta.data
  tibble(
    sample = sample_id,
    n_cells = nrow(md),
    n_singlet = sum(md$scDblFinder.class == "singlet", na.rm = TRUE),
    n_doublet = sum(md$scDblFinder.class == "doublet", na.rm = TRUE)
  )
})) %>%
  mutate(doublet_rate = round(100 * n_doublet / n_cells, 2))

write.csv(doublet_summary, file.path(out_dir, "doublet_summary.csv"), row.names = FALSE)
doublet_summary
save_doublet_plots(seurat_objs, out_dir, doublet_summary = doublet_summary)
```

---

## Step 3 — Merge

```{r step3-merge, eval=run_step1_3_build_merge}
seurat_all <- merge(
  x = seurat_objs[[1]],
  y = seurat_objs[-1],
  add.cell.ids = names(seurat_objs),
  project = merge_project
)
table(seurat_all$orig.ident)
saveRDS(seurat_all, merged_raw_rds)
rm(seurat_objs)
gc()
```

```{r step3-load-raw, eval=!run_step1_3_build_merge}
seurat_all <- readRDS(merged_raw_rds)
```

---

## Step 4 — Exploratory QC

```{r step4-compute-thresholds, eval=run_step4_qc_exploratory}
seurat_all <- readRDS(merged_raw_rds)
seurat_all[["percent.mt"]] <- PercentageFeatureSet(seurat_all, pattern = "^mt-")

md <- seurat_all@meta.data
md$orig.ident <- factor(md$orig.ident, levels = sort(unique(md$orig.ident)))

per_sample_thresholds <- compute_per_sample_nfeature_max(
  md,
  nFeature_min = nFeature_min,
  sd_multiplier = sd_multiplier
)
cell_filter <- make_median5sd_cell_filter(
  mt_max = mt_max,
  per_sample_thresholds = per_sample_thresholds,
  nFeature_min = nFeature_min,
  sd_multiplier = sd_multiplier
)

per_sample_thresholds
threshold_folder_label_median5sd(cell_filter)
```

```{r step4-qc-plots, eval=run_step4_qc_exploratory}
plot_qc_density_per_sample_max(
  md, "nFeature_RNA", "nFeature — density (zoom)", "nFeature_RNA",
  per_sample_max = cell_filter$per_sample_max,
  vline_min = cell_filter$nFeature_min,
  xlim = density_xlim_nFeature
)
plot_qc_density(md, "nCount_RNA", "nCount — density (zoom)", "nCount_RNA",
                NULL, NULL, xlim = density_xlim_nCount)
plot_qc_box_per_sample_max(
  md, "nFeature_RNA", "nFeature — boxplot", "nFeature_RNA",
  per_sample_max = cell_filter$per_sample_max,
  hline_min = cell_filter$nFeature_min
)
plot_qc_box(md, "nCount_RNA", "nCount — boxplot", "nCount_RNA", NULL, NULL)
plot_qc_scatter_per_sample_max(
  md, "nFeature vs nCount — scatter",
  per_sample_max = cell_filter$per_sample_max,
  nFeature_min = cell_filter$nFeature_min
)

save_qc_exploratory_plots_median5sd(
  md, cell_filter, out_dir, gene_label = gene_label,
  density_xlim_nFeature = density_xlim_nFeature,
  density_xlim_nCount = density_xlim_nCount
)

if (!is.null(qc_zoom_y_min) && !is.null(qc_zoom_y_max)) {
  save_qc_zoom_box_plots_median5sd(
    md, out_dir, gene_label = gene_label,
    th = cell_filter,
    y_min = qc_zoom_y_min, y_max = qc_zoom_y_max
  )
} else {
  message("Skipping zoom boxplots (set qc_zoom_y_min and qc_zoom_y_max to enable).")
}
```

---

## Step 4a — nFeature summary tables

```{r step4a-threshold-diagnostics, eval=run_step4a_filter_diagnostics}
if (!exists("cell_filter")) {
  seurat_all <- readRDS(merged_raw_rds)
  if (!"percent.mt" %in% colnames(seurat_all@meta.data)) {
    seurat_all[["percent.mt"]] <- PercentageFeatureSet(seurat_all, pattern = "^mt-")
  }
  md_qc <- seurat_all@meta.data
  md_qc$orig.ident <- factor(md_qc$orig.ident, levels = sort(unique(md_qc$orig.ident)))
  per_sample_thresholds <- compute_per_sample_nfeature_max(
    md_qc, nFeature_min = nFeature_min, sd_multiplier = sd_multiplier
  )
  cell_filter <- make_median5sd_cell_filter(
    mt_max = mt_max,
    per_sample_thresholds = per_sample_thresholds,
    nFeature_min = nFeature_min,
    sd_multiplier = sd_multiplier
  )
} else {
  md_qc <- md
}

threshold_diagnostics <- run_median5sd_threshold_diagnostics(
  md = md_qc,
  out_dir = out_dir,
  cell_filter = cell_filter
)

threshold_diagnostics$distribution
threshold_diagnostics$per_sample_max
```

---

## Step 4b — SD/MAD bounds and round-number grid

```{r step4b-threshold-tables, eval=run_step4b_round_grid_preview}
if (!exists("threshold_diagnostics")) {
  stop("Run Step 4a first.")
}
threshold_diagnostics$sd_mad_bounds[
  threshold_diagnostics$sd_mad_bounds$scope == "overall",
]
threshold_diagnostics$round_grid
```

---

## Step 5 — Filter

```{r step5-filter, eval=run_step5_filter}
seurat_all <- readRDS(merged_raw_rds)
seurat_all[["percent.mt"]] <- PercentageFeatureSet(seurat_all, pattern = "^mt-")

if (!exists("cell_filter")) {
  md_tmp <- seurat_all@meta.data
  per_sample_thresholds <- compute_per_sample_nfeature_max(
    md_tmp, nFeature_min = nFeature_min, sd_multiplier = sd_multiplier
  )
  cell_filter <- make_median5sd_cell_filter(
    mt_max = mt_max,
    per_sample_thresholds = per_sample_thresholds,
    nFeature_min = nFeature_min,
    sd_multiplier = sd_multiplier
  )
}

md_before <- seurat_all@meta.data
trial <- run_median5sd_threshold_trial(
  seurat_all,
  cell_filter,
  out_dir,
  md_before = md_before,
  density_xlim_nFeature = density_xlim_nFeature,
  density_xlim_nCount = density_xlim_nCount,
  qc_zoom_y_min = if (!is.null(qc_zoom_y_min)) qc_zoom_y_min else 200,
  qc_zoom_y_max = if (!is.null(qc_zoom_y_max)) qc_zoom_y_max else 3000
)
trial$filter_summary$overall
trial$filter_summary$cells_per_sample

saveRDS(filter_seurat_by_median5sd(seurat_all, cell_filter), merged_filt_rds)
```

---

## Step 6 setup (shared)

```{r step6-setup, eval=run_step6a_sctransform || run_step6b_clustree || run_step6c_clustering}
snRNAseq_median5sd_clustree_init(repo_root = repo_root)

if (!exists("cell_filter")) {
  cell_filter <- build_cell_filter_from_raw(
    merged_raw_rds,
    nFeature_min = nFeature_min,
    sd_multiplier = sd_multiplier,
    mt_max = mt_max
  )
}

step6_cl_dir <- clustering_dir_median5sd(out_dir, cell_filter)
step6_paths <- clustering_rds_paths(step6_cl_dir, object_prefix)
threshold_folder_label_median5sd(cell_filter)
step6_paths
```

---

## Step 6a — SCTransform, Harmony, UMAP, FindNeighbors

Merged filtered object; **LogNormalize fallback** if SCT fails.

```{r step6a-sctransform, eval=run_step6a_sctransform}
step6a_result <- run_step6a_sctransform(
  raw_rds = merged_raw_rds,
  th = cell_filter,
  out_dir = out_dir,
  object_prefix = object_prefix,
  force = step6_force_rerun
)
step6a_result
if (file.exists(file.path(step6_cl_dir, "sctransform_summary.csv"))) {
  read.csv(file.path(step6_cl_dir, "sctransform_summary.csv"), stringsAsFactors = FALSE)
}
```

---

## Step 6b — clustree

Requires Step 6a. Review `clustree.png`, set `clustering_resolutions`, then enable Step 6c.

```{r step6b-clustree, eval=run_step6b_clustree}
step6b_result <- run_step6b_clustree(
  th = cell_filter,
  out_dir = out_dir,
  object_prefix = object_prefix,
  clustree_resolutions = clustree_resolutions,
  force = step6_force_rerun
)
step6b_result
if (file.exists(file.path(step6_cl_dir, "clustree_resolution_columns.csv"))) {
  read.csv(file.path(step6_cl_dir, "clustree_resolution_columns.csv"), stringsAsFactors = FALSE)
}
```

---

## Step 6c — Clustering + UMAP plots

Edit **`clustering_resolutions`** before running.

```{r step6c-clustering, eval=run_step6c_clustering}
clustering_summary <- run_step6c_clustering(
  th = cell_filter,
  out_dir = out_dir,
  object_prefix = object_prefix,
  resolutions = clustering_resolutions,
  force = step6_force_rerun
)
clustering_summary
```

---

## Optional — DESeq2

```r
# C:/Yale/R/deseq2_snrna_pseudobulk_export.R
```

---

## Output index

| Path | Step |
|------|------|
| `dyrk1a_merged_raw.rds` | 3 |
| `qc_plots/` | 4 |
| `threshold_diagnostics/` | 4a–4b |
| `qc_filter_plots/<label>/` | 5 |
| `dyrk1a_merged_filtered_median5sd.rds` | 5 |
| `clustering/<label>/dyrk1a_harmony_prepped.rds`, `sctransform_summary.csv` | 6a |
| `clustering/<label>/clustree.png`, `dyrk1a_clustree.rds` | 6b |
| `clustering/<label>/umap_*.png`, `dyrk1a_harmony_res*.rds` | 6c |
| `<label>` = `nFmin200_nFmax_median5SD_mtmax3` | |
